Make any stupid plot you want

### 2021-05-06 draft/sections/math.tex
artist equivalence

\begin{figure}[htb]
  \centering
  \begin{subfigure}{.49\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
    \caption{}
    \label{fig:math:artist:nu:valid}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/partial_invalid.png}
    \caption{}
    \label{fig:math:artist:nu:invalid}
  \end{subfigure}
  \caption{ The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} represented by the colored arrows is monotonic, and therefore monoid equivariant,  since $\nu(storm)=\nu(rain)$ satisfies the condition $\nu(storm) \geq \nu(storm)$. In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because $rain \geq storm$ is mapped to elements with the reverse ordering $\nu(storm) \geq \nu(storm)$.}
  \label{fig:math:artist:nu}
\end{figure}
The mapping from weather state to umbrellla in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$. \autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms. In contrast, the translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not monotonic and therefore not equivariant This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\). To satisfy the monotonic condition for $rain \geq storm$, either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.

@@

\label{sec:math:artist:q}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
  \caption{The transform functions $\vchannel_i$ convert data $\dsection_i\in \dtotal$ to visual characteristics $\vsection_i \in \vtotal$. These visual components $\vsection_i$ are then assembled by \vmark\ into a graphic $\gsection \in \gtotal$.} 
  \label{fig:math:artist:q}
\end{figure}
The transformation from data into graphic is analagous to a map-reduce operation; as illustrated in \autoref{fig:math:artist:q}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal. The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$. The full space of graphics is not necessarily equivariant; therefore we formalize the constraints on \vmark\ such that it produces structure preserving graphics.

We formalize the expectation that visualization generation functions parameterized in the same way should generate the same functions as the equivariant map $\vmark: \vsection \mapsto \gsection$. We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.  We do not define monoid actions on all of $\Gamma(\gtotal)$ because there may be graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
\begin{figure}[htb]
  \centering
  \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
  \caption{These two glyphs are generated by the same annulus \vmark\ function. The monoid action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
  \label{fig:math:artist:graphic}
\end{figure}
Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$
\begin{prop}
If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
\begin{equation}
\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
\label{eq:math:artist:q}
\end{equation}
then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$. If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection. 
\end{prop}

@@

\subsubsection{Assembly Template\vmarkd}
\label{sec:math:artist:qhat}
The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark. As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
%[column sep=huge]
\begin{equation}
    \begin{tikzcd}
        \textcolor{gray!50}{\dtotal} \arrow[r, "\vchannel", color=gray!50] \arrow[rd, "\pi"', color=gray!50] & \vtotal \arrow[d, "\pi"']                  & \vindex^*\vtotal \arrow[r, "\vmark", color=gray!50] \arrow[d, "\vindex^*\pi"'] \arrow[l,  "\vindex^*"'] & \textcolor{gray!50}{\gtotal} \arrow[ld, "\pi", color=gray!50] \\
                                                          & \dbase \arrow[u, "\vsection"', bend right] & \gbase \arrow[l, "\vindex"'] \arrow[u, "\vindex^*\vsection"', bend right]               &                          
        \end{tikzcd}
        \label{eq:math:artist:qhat}
\end{equation}

@@

\subsubsection{Composition of Artists: +}
\begin{figure}[H]
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
    \caption{}
    \label{fig:math:combined}    
\end{subfigure}
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
    \caption{}
    \label{fig:math:seperate}
\end{subfigure}
\caption{ In \autoref{fig:math:combined}, these artists are composited before being added to the image. Disjoint union of \dtotal\ aligns the two timeseries with the x and y axis so all these elements use a shared coordinate system. A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying. None of this machinary exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independented of the other elements.}
\label{fig:artist_plus}
\end{figure}

We describe the constraints for compositing artists by defining addition operators. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 
\begin{equation}
+ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
\end{equation}
defines a simple composition of artists. For example, the components in \autoref{fig:math:seperate} are each generated by different artists, and a visualization of solely the x axis is rarely all that useful. In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on how they are generated in conjunction with each other. In \autoref{fig:math:combined}, the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system. When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section. This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.

\subsubsection{Equivalance class of artists $\vartist^{\prime}$}
\label{sec:artist_equivalance}
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/equivalent_artists.png}
   \caption{Each scatter plot is generated via a unique artist function $\vartist_{i}$, but they only differ in aesthetic styling. Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartist{\prime}$}
    \label{fig:math:artist:equivalence}
\end{figure}
Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization\cite{kindlmannAlgebraicProcessVisualization2014}. We propose that visualizations are invariant if they are generated by artists that are members of an equivalence class 
\begin{equation*}
\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}
\label{eq:math:artist:aprime}
\end{equation*}
For example, every scatter plot in \autoref{fig:math:artist:equivalence} is a scatter of the same datasets mapped to the \textit{x position} and \textit{y position} in the same way. The scatter plots only differ in the choice of constant visual literals, differing in color and marker shape. Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartist{\prime}$. Since it is impractical to implement a new artist for every single graphic, the equivalance class provides a way to evaluate an implementation of a generalized artist. Given equivalent, but no necessarily identical, \vchannel, \vmark, and \vindex, two artists are equivalent. This criteria also allows for comparing artists across libraries. 
\end{document}

>>>>

\begin{figure}[htb]
  \centering
  \begin{subfigure}{.49\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
    \caption{}
    \label{fig:math:artist:nu:valid}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/partial_invalid.png}
    \caption{}
    \label{fig:math:artist:nu:invalid}
  \end{subfigure}
  \caption{ The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} represented by the colored arrows is monotonic, and therefore monoid equivariant,  since $\nu(storm)=\nu(rain)$ satisfies the condition $\nu(storm) \geq \nu(storm)$. In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because $rain \geq storm$ is mapped to elements with the reverse ordering $\nu(storm) \geq \nu(storm)$.}
  \label{fig:math:artist:nu}
\end{figure}
The mapping from weather state to umbrella in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$. \autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms. In contrast, the translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not monotonic and therefore not equivariant This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\). To satisfy the monotonic condition for $rain \geq storm$, either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.

@@

\label{sec:math:artist:q}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
  \caption{The transform functions $\vchannel_i$ convert data $\dsection_i\in \dtotal$ to visual characteristics $\vsection_i \in \vtotal$. These visual components $\vsection_i$ are then assembled by \vmark\ into a graphic $\gsection \in \gtotal$.} 
  \label{fig:math:artist:q}
\end{figure}
The transformation from data into graphic is analogous to a map-reduce operation; as illustrated in \autoref{fig:math:artist:q}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal. The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$. The full space of graphics is not necessarily equivariant; therefore we formalize the constraints on \vmark\ such that it produces structure preserving graphics.

We define the visualization assembly function $\vmark: \vsection \mapsto \gsection$ as an equivariant map to formalize the expectation that two \vmark\ functions parameterized in the same way should generate the same graphic. We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.  We do not define monoid actions on all of $\Gamma(\gtotal)$ because there may be graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
\begin{figure}[htb]
  \centering
  \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
  \caption{These two glyphs are generated by the same annulus \vmark\ function. The monoid action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
  \label{fig:math:artist:graphic}
\end{figure}
Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$
\begin{prop}
If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
\begin{equation}
\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
\label{eq:math:artist:q}
\end{equation}
then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$. If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection. 
\end{prop}

@@

\subsubsection{Assembly Template \vmarkd}
\label{sec:math:artist:qhat}
The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark. As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
%[column sep=huge]
\begin{equation}
    \begin{tikzcd}
        \textcolor{gray!50}{\dtotal} \arrow[r, "\vchannel", color=gray!50] \arrow[rd, "\pi"', color=gray!50] & \vtotal \arrow[d, "\pi"']                  & \vindex^*\vtotal \arrow[r, "\vmark", color=gray!50] \arrow[d, "\vindex^*\pi"'] \arrow[l,  "\vindex^*"'] & \textcolor{gray!50}{\gtotal} \arrow[ld, "\pi", color=gray!50] \\
                                                          & \dbase \arrow[u, "\vsection"', bend right] & \gbase \arrow[l, "\vindex"'] \arrow[u, "\vindex^*\vsection"', bend right]               &                          
        \end{tikzcd}
        \label{eq:math:artist:qhat}
\end{equation}

@@

\subsubsection{Composition of Artists: +}
\begin{figure}[H]
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
    \caption{}
    \label{fig:math:combined}    
\end{subfigure}
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
    \caption{}
    \label{fig:math:seperate}
\end{subfigure}
\caption{ In \autoref{fig:math:combined}, these artists are composited before being added to the image. Disjoint union of \dtotal\ aligns the two timeseries with the x and y axis so all these elements use a shared coordinate system. A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying. None of this machinary exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independented of the other elements.}
\label{fig:artist_plus}
\end{figure}
Visualizations generally consist of more than one artist, commonly having visual elements such as the plot and axis labels and maybe legends. To generate these composite images, we define addition operators and specify the constraints for compositing artists. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 
\begin{equation}
+ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
\end{equation}
defines a simple composition of artists. For example, in \autoref{fig:math:combined} the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system. In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on where they are placed in the image. When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section. This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.

\subsubsection{Equivalance class of artists $\vartist^{\prime}$}
\label{sec:artist_equivalance}
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/equivalent_artists.png}
   \caption{Each scatter plot is generated via a unique artist function $\vartist_{i}$, but they only differ in aesthetic styling. Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartisteq$}
    \label{fig:math:artist:equivalence}
\end{figure}
Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization\cite{kindlmannAlgebraicProcessVisualization2014}. By defining a criteria for invariance, we can evaluate whether two artists generate the same type of graphic and compare artists across libraries. We propose that visualizations are invariant if they are generated by artists that are members of an equivalence class 
\begin{equation*}
\{\vartist \in \vartisteq: \vartist_{1} \equiv \vartist_{2}\}
\label{eq:math:artist:aprime}
\end{equation*}
For example, every scatter plot in \autoref{fig:math:artist:equivalence} is a scatter of the same datasets mapped to the \textit{x position} and \textit{y position} in the same way. The scatter plots only differ in the choice of constant visual literals, differing in color and marker shape. Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartisteq$. Since it is impractical to implement a new artist for every single graphic, the equivalance class provides a way to evaluate an implementation of a generalized artist. 
\end{document}
